A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-23 tex/MercuryNumerical.tex
Group edditing session applied

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed that allows one to estimate the size of the effect as well as the uncertainty of 
the approach a priori. At the same time this enables the students to double check the results found numerically.
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students familiar to derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect as well as the 
expected accuracy of a calculation, but also to cross check whether the results of the simulation are sensible.
As is pointed out in the chapters below the course can be finished at some canonical points --- in particular we think
that for most students a detailed study of the uncertainties of the numerical simulations might be too technical.
However, we put also this discussion into the paper not only for completeness but also to provide an additional 
challenge to the more advanced students.

@@

This procedure can only work, if $\Delta t$ is sufficiently small.
What this means depends on the process one studies.
One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly. This issue will
be discussed below in more detail.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General Relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88.0\,T_0$.
With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
\begin{eqnarray}
r_{MS}(0)   &=& 4.60 R_0 \, , \quad
v_{M}(0)    = 0.510 \frac{R_0}{T_0} \, ,  \\
a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
\end{eqnarray}

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
TM  = 8.80e+1 # Orbit period of Mercury
rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
\end{lstlisting}
where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.

It is handy to use \texttt{Python}s functions to structure the program and hold repeating code ("DRY" --- Don't Repeat Yourself):
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
    # Compute the strength of the acceleration
    temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
    aMS  = c_a * temp / vec_rM_old.mag**2
    # Multiply by the direction
    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
    # Update velocity vector
    vec_vM_new = vec_vM_old + vec_aMS * dt
    # Update position vector
    vec_rM_new = vec_rM_old + vec_vM_new * dt
    return vec_rM_new, vec_vM_new

@@

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to \texttt{import} further objects from the \texttt{VPython} module.
We change the import statement from before to
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent Mercury and the Sun in the simulation
\begin{lstlisting}
# Define the initial coordinates; M = Mercury, S = Sun
M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# And the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# Add a visible trajectory to Mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We place the Sun in the origin of our coordinate system and choose non-realistic radii for visualization purposes.
Last but not least, one should the vectors related to the visualization directly instead of extra vectors as was done before. Thus
the \texttt{while}-loop becomes
\begin{lstlisting}
t     = 0.0
alpha = 0.0
beta  = 0.0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
    # Set the frame rate (you can choose a higher rate to accelerate the program)
    rate(100)
    # Update the drawn trajectory with the current position
    M.trajectory.append(pos=M.pos)
    # Update velocity and position
    M.pos, M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
    # Advance time by one step
    t = t + dt
\end{lstlisting}

>>>>

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed that allows one to estimate the size of the effect as well as the uncertainty of 
the approach a priori. At the same time this enables the students to double check the results found numerically.
The course is structured into a basic block and two further refinements which aim at more advanced students.
\end{abstract}

@@

Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition, tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool to estimate the size of the effect studied as well as its expected accuracy.
Furthermore the method allows one to cross check the results of the simulation.
As is pointed out in the chapters below the course can be finished at some canonical points --- in particular we think
that for most students a detailed study of the uncertainties of the numerical simulations might be too technical.
However, we include this discussion as well both for completeness and as an additional challenge to the more advanced students.

@@

This procedure can only work if $\Delta t$ is sufficiently small.
One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive to the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
This issue will be discussed in more detail below.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable \cite{Einstein}
\begin{equation}
\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT} \, .
\end{equation}

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha \, \frac{r_S}{r} )$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_L^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_L^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta \, \frac{r_L^2}{r^2}$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_L^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_L^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{ r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other (see Figure \ref{fig:sun_merc}).
We thus find that $(r_L^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore one may expect from both correction terms an effect of similar size --- for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}.
Furthermore, any term of higher order in either $r_L^2$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General Relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59.0\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88.0\,T_0$.
With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
\begin{eqnarray}
r_{MS}(0)   &=& 4.60 R_0 \, , \quad
v_{M}(0)    = 0.510 \frac{R_0}{T_0} \, ,  \\
a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r_{MS}^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
\end{eqnarray}

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.60    # Initial radius of Mercury orbit, in units of R0
vM0 = 5.10e-1 # Initial orbital speed of Mercury, in units of R0/T0
c_a = 9.90e-1 # Base acceleration of Mercury, in units of R0**3/T0**2
TM  = 8.80e+1 # Orbit period of Mercury
rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
rL2 = 8.19e-7 # Specific angular momentum, in units of R0**2
\end{lstlisting}
where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_L^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.

So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpython
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.

It is handy to use \texttt{Python}s functions to structure the program and hold repeating code ("DRY" --- Don't Repeat Yourself):
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
    # Compute the strength of the acceleration
    temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
    aMS  = c_a * temp / vec_rM_old.mag**2
    # Multiply by the direction
    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
    # Update velocity vector
    vec_vM_new = vec_vM_old + vec_aMS * dt
    # Update position vector
    vec_rM_new = vec_rM_old + vec_vM_new * dt
    return vec_rM_new, vec_vM_new

@@

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to \texttt{import} further objects from the \texttt{VPython} module.
We change the import statement from before to
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent Mercury and the Sun in the simulation
\begin{lstlisting}
# Define the initial coordinates; M = Mercury, S = Sun
M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# And the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# Add a visible trajectory to Mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We place the Sun in the origin of our coordinate system and choose non-realistic radii for visualization purposes.
Last but not least, one should use in the code the vectors directly related to the visualization. Thus
the \texttt{while}-loop becomes
\begin{lstlisting}
t     = 0.0
alpha = 0.0
beta  = 0.0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
    # Set the frame rate (you can choose a higher rate to accelerate the program)
    rate(100)
    # Update the drawn trajectory with the current position
    M.trajectory.append(pos=M.pos)
    # Update velocity and position
    M.pos, M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
    # Advance time by one step
    t = t + dt
\end{lstlisting}

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
[Figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}
